---
title: "PBMC10X RNA Analysis"
author: "Siyuan Luo"
date: "2022-09-26"
output:
  html_document:
    df_print: paged
    toc: TRUE
    toc_float: TRUE
    theme: default
# editor_options:
#   chunk_output_type: console
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, eval=FALSE}
readRenviron("~/.Renviron")
library("reticulate")
py_config()
```

```{r,eval=FALSE}
# install
setRepositories(ind=1:3) # needed to automatically install Bioconductor dependencies
install.packages("Signac")
BiocManager::install(c('BSgenome.Hsapiens.UCSC.hg19', 'EnsDb.Hsapiens.v75'))
BiocManager::install(c('BSgenome.Hsapiens.UCSC.hg38', 'EnsDb.Hsapiens.v86'))
BiocManager::install(c('BSgenome.Mmusculus.UCSC.mm10', 'EnsDb.Mmusculus.v79'))
```

```{r,message=FALSE}
library(Seurat)
library(GenomeInfoDb)
library(EnsDb.Hsapiens.v86)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ggplot2)
library(patchwork)
library(Matrix)
library(dplyr)
library(scDblFinder)
library(BiocSingular)
library(scater)

dyn.load('/usr/local/hdf5/lib/libhdf5.so.200')
library(hdf5r)
set.seed(1234)

```

```{r}
# computer specific settings
proj_dir <- "/home/siluo/nme_public/SiyuanLuo/projects/sc_chromatin_benchmark/"
software_dir <- "/home/siluo/Software/"
setwd(paste0(proj_dir, "PBMC_multi-omics_10X/rna"))
```

```{r}
pbmc_rna <- readRDS("pbmc_rna_preprocessed.rds")
```

# Redo clustering and annotation

## Redo clustering after doublet removal
```{r, fig.keep='all',message=FALSE,warning = FALSE, results='hide'}
pbmc_rna <- SCTransform(pbmc_rna)
pbmc_rna <- RunPCA(pbmc_rna, features = VariableFeatures(object = pbmc_rna))
ElbowPlot(pbmc_rna)
pbmc_rna <- FindNeighbors(pbmc_rna, 
                          dims = 1:11,
                          graph.name = c("SCT_nn_PCAn11","SCT_snnPCAn11"))

pbmc_rna <- FindNeighbors(pbmc_rna, 
                          dims = 1:20,
                          graph.name = c("SCT_nn_PCAn20","SCT_snnPCAn20"))

pbmc_rna <- FindClusters(pbmc_rna, resolution = 0.8, graph.name = "SCT_snnPCAn11", algorithm = 4)

pbmc_rna <- FindClusters(pbmc_rna, resolution = 0.8, graph.name = "SCT_snnPCAn20", algorithm = 4)

pbmc_rna <- RunUMAP(pbmc_rna, dims = 1:11)
pbmc_rna$rna_clusters_r0.8 <- pbmc_rna$SCT_snnPCAn11_res.0.8
DimPlot(pbmc_rna, reduction = "umap", group.by="SCT_snnPCAn11_res.0.8",cols="polychrome")
DimPlot(pbmc_rna, reduction = "umap", group.by="SCT_snnPCAn20_res.0.8",cols="polychrome")
```

## Cross checking
```{r}
## Check the consensus between clustering and label transfer
df_count <- dplyr::rename(dplyr::count(pbmc_rna[[c("rna_clusters_r0.8","predicted.id")]], rna_clusters_r0.8, predicted.id), Freq = n) 
df_count <- dplyr::arrange(df_count, rna_clusters_r0.8, -Freq)
df_count
```
## Subset and recluster {.tabset}

### CD14 Mono

#### UMAP
```{r, fig.keep='all',message=FALSE,warning = FALSE, results='hide'}
pbmc_sub <- subset(x=pbmc_rna, subset=rna_clusters_r0.8 %in% c(1,4,7,8))
pbmc_sub <- SCTransform(pbmc_sub)
pbmc_sub <- RunPCA(pbmc_sub, features = VariableFeatures(object = pbmc_sub))
ElbowPlot(pbmc_sub)
pbmc_sub <- FindNeighbors(pbmc_sub, dims = 1:7)
pbmc_sub <- FindClusters(pbmc_sub, resolution = 0.3)
pbmc_sub <- RunUMAP(pbmc_sub, dims = 1:7)

DimPlot(pbmc_sub, reduction = "umap",group.by="rna_clusters_r0.8",cols="polychrome")
DimPlot(pbmc_sub, reduction = "umap",group.by="predicted.id",cols="polychrome")
DimPlot(pbmc_sub, reduction = "umap",cols="polychrome")
FeaturePlot(pbmc_sub,features = "nCount_RNA")
FeaturePlot(pbmc_sub,features = "nFeature_RNA")
FeaturePlot(pbmc_sub,features = "percent.mt")
```

#### Marker
```{r,message=FALSE,warning = FALSE}
markers <- FindAllMarkers(pbmc_sub)
top20 <- dplyr::top_n(dplyr::group_by(markers, cluster), n = 20, wt = avg_log2FC)

top20 # the fold change is large, but not actually DE ??? 
DoHeatmap(pbmc_sub, features = markers$gene)
```

#### Silhouette
```{r,message=FALSE,warning = FALSE}
pbmc_sub$no_cluster <- 1
pbmc_sub$no_cluster[pbmc_sub$rna_clusters_r0.8==8] <- 2
pca <- Embeddings(Reductions(pbmc_sub,"pca"))
dist.matrix <- dist(x = pca[, 1:7])
clusters1 <- pbmc_sub$rna_clusters_r0.8
clusters2 <- pbmc_sub$no_cluster
sil1 <- cluster::silhouette(x = as.numeric(x = as.factor(x = clusters1)), dist = dist.matrix)
sil2 <- cluster::silhouette(x = as.numeric(x = as.factor(x = clusters2)), dist = dist.matrix)
mean(sil1[, 3])
mean(sil2[, 3])
```

Merge cluster 1,4,7 together

### CD4/CD8 Naive

#### Marker
```{r,message=FALSE,warning = FALSE}
pbmc_sub <- subset(x=pbmc_rna, subset=rna_clusters_r0.8 %in% c(2,3,11,16,5,6))
markers <- FindAllMarkers(pbmc_sub)
top20 <- dplyr::top_n(dplyr::group_by(markers, cluster), n = 20, wt = avg_log2FC)

top20 # the fold change is large, but not actually DE ??? 
DoHeatmap(pbmc_sub, features = markers$gene)
```

#### UMAP
```{r, fig.keep='all',message=FALSE,warning = FALSE, results='hide'}
pbmc_sub <- SCTransform(pbmc_sub)
pbmc_sub <- RunPCA(pbmc_sub, features = VariableFeatures(object = pbmc_sub))
ElbowPlot(pbmc_sub)
pbmc_sub <- FindNeighbors(pbmc_sub, dims = 1:10)
pbmc_sub <- FindClusters(pbmc_sub, resolution = 0.3)
pbmc_sub <- RunUMAP(pbmc_sub, dims = 1:10)

DimPlot(pbmc_sub, reduction = "umap",group.by="rna_clusters_r0.8",cols="polychrome")
DimPlot(pbmc_sub, reduction = "umap",group.by="predicted.id",cols="polychrome")
DimPlot(pbmc_sub, reduction = "umap",cols="polychrome")
FeaturePlot(pbmc_sub,features = "nCount_RNA")
FeaturePlot(pbmc_sub,features = "nFeature_RNA")
FeaturePlot(pbmc_sub,features = "percent.mt")
```
Keep the seperation between 2 and 16, 3 and 11


### CD4 TCM

#### Marker
```{r,message=FALSE,warning = FALSE}
pbmc_sub <- subset(x=pbmc_rna, subset=rna_clusters_r0.8 %in% c(5, 6, 13))
markers <- FindAllMarkers(pbmc_sub)
top20 <- dplyr::top_n(dplyr::group_by(markers, cluster), n = 20, wt = avg_log2FC)

top20
DoHeatmap(pbmc_sub, features = markers$gene)
```

#### UMAP
```{r, fig.keep='all',message=FALSE,warning = FALSE, results='hide'}
pbmc_sub <- SCTransform(pbmc_sub)
pbmc_sub <- RunPCA(pbmc_sub, features = VariableFeatures(object = pbmc_sub))
ElbowPlot(pbmc_sub)
pbmc_sub <- FindNeighbors(pbmc_sub, dims = 1:6)
pbmc_sub <- FindClusters(pbmc_sub, resolution = 0.2)
pbmc_sub <- RunUMAP(pbmc_sub, dims = 1:6)

DimPlot(pbmc_sub, reduction = "umap",group.by="rna_clusters_r0.8",cols="polychrome")
DimPlot(pbmc_sub, reduction = "umap",group.by="predicted.id",cols="polychrome")
DimPlot(pbmc_sub, reduction = "umap",cols="polychrome")
FeaturePlot(pbmc_sub,features = "nCount_RNA")
FeaturePlot(pbmc_sub,features = "nFeature_RNA")
FeaturePlot(pbmc_sub,features = "percent.mt")
```

#### Silhouette
```{r,message=FALSE,warning = FALSE}
pbmc_sub$no_cluster <- pbmc_sub$rna_clusters_r0.8
pbmc_sub$no_cluster[pbmc_sub$rna_clusters_r0.8==6] <- 3
# pbmc_sub$no_cluster[pbmc_sub$seurat_clusters==3] <- 4
table(pbmc_sub$no_cluster)
pca <- Embeddings(Reductions(pbmc_sub,"pca"))
dist.matrix <- dist(x = pca[, 1:6])
clusters1 <- pbmc_sub$rna_clusters_r0.8
clusters2 <- pbmc_sub$no_cluster
sil1 <- cluster::silhouette(x = as.numeric(x = as.factor(x = clusters1)), dist = dist.matrix)
sil2 <- cluster::silhouette(x = as.numeric(x = as.factor(x = clusters2)), dist = dist.matrix)
mean(sil1[, 3])
mean(sil2[, 3])
```
The seperation between 5 and 6 seems to be very random. The marker of cluster 6 include a lot of ribosomal protein genes, which may suggest some technical effect. Merge cluster 5 and 6.

### B cell subtypes

#### UMAP
```{r, fig.keep='all',message=FALSE,warning = FALSE, results='hide'}
pbmc_sub <- subset(x=pbmc_rna, subset=rna_clusters_r0.8 == 10)
pbmc_sub <- SCTransform(pbmc_sub)
pbmc_sub <- RunPCA(pbmc_sub, features = VariableFeatures(object = pbmc_sub))
ElbowPlot(pbmc_sub)
pbmc_sub <- FindNeighbors(pbmc_sub, dims = 1:10)
pbmc_sub <- FindClusters(pbmc_sub, resolution = 0.3)
pbmc_sub <- RunUMAP(pbmc_sub, dims = 1:10)

DimPlot(pbmc_sub, reduction = "umap",group.by="rna_clusters_r0.8",cols="polychrome")
DimPlot(pbmc_sub, reduction = "umap",group.by="predicted.id",cols="polychrome")
DimPlot(pbmc_sub, reduction = "umap",cols="polychrome")
FeaturePlot(pbmc_sub,features = "nCount_RNA")
FeaturePlot(pbmc_sub,features = "nFeature_RNA")
FeaturePlot(pbmc_sub,features = "percent.mt")
```

#### Marker
```{r,message=FALSE,warning = FALSE}
markers <- FindAllMarkers(pbmc_sub)
top20 <- dplyr::top_n(dplyr::group_by(markers, cluster), n = 20, wt = avg_log2FC)

top20
DoHeatmap(pbmc_sub, features = markers$gene)
```

In cluster 10, intermediate B cells and memory B cells are not separated by the clustering algorithm. The markers above show 2 marker of intermediate B cell and 3 marker of memory B cell. UMAP show that there're indeed two populations there. Within cluster 10, redo clustering, to give two populations different labels.

```{r}
cell_inter <- Cells(subset(x=pbmc_sub, subset=rna_clusters_r0.8 == 10 & pbmc_sub$seurat_clusters == 0))
cell_memory <- Cells(subset(x=pbmc_sub, subset=rna_clusters_r0.8 == 10 & pbmc_sub$seurat_clusters == 1))
```

### Cluster 12, 13

#### UMAP
```{r, fig.keep='all',message=FALSE,warning = FALSE, results='hide'}
pbmc_sub <- subset(x=pbmc_rna, subset=rna_clusters_r0.8 %in% c(12,13))
pbmc_sub <- SCTransform(pbmc_sub)
pbmc_sub <- RunPCA(pbmc_sub, features = VariableFeatures(object = pbmc_sub))
ElbowPlot(pbmc_sub)
pbmc_sub <- FindNeighbors(pbmc_sub, dims = 1:10)
pbmc_sub <- FindClusters(pbmc_sub, resolution = 0.2)
pbmc_sub <- RunUMAP(pbmc_sub, dims = 1:10)

DimPlot(pbmc_sub, reduction = "umap",group.by="rna_clusters_r0.8",cols="polychrome")
DimPlot(pbmc_sub, reduction = "umap",group.by="predicted.id",cols="polychrome")
DimPlot(pbmc_sub, reduction = "umap",cols="polychrome")
FeaturePlot(pbmc_sub,features = "nCount_RNA")
FeaturePlot(pbmc_sub,features = "nFeature_RNA")
FeaturePlot(pbmc_sub,features = "percent.mt")
```

#### Marker
```{r,message=FALSE,warning = FALSE}
markers <- FindAllMarkers(pbmc_sub)
top20 <- dplyr::top_n(dplyr::group_by(markers, cluster), n = 20, wt = avg_log2FC)

top20
DoHeatmap(pbmc_sub, features = markers$gene)
```
Seperate cluster 11 into 2 clsuters, seperate cluster 12 into 2 clsuters.

```{r}
cell_cd8tem1 <- Cells(subset(x=pbmc_sub, subset=rna_clusters_r0.8 == 12 & pbmc_sub$seurat_clusters == 1))
cell_mait <- Cells(subset(x=pbmc_sub, subset=rna_clusters_r0.8 == 12 & pbmc_sub$seurat_clusters == 2))
cell_cd8tem2 <- Cells(subset(x=pbmc_sub, subset=rna_clusters_r0.8 == 13 & pbmc_sub$seurat_clusters == 0))
cell_gdt <- Cells(subset(x=pbmc_sub, subset=rna_clusters_r0.8 == 13 & pbmc_sub$seurat_clusters == 3))
```

## Final labels 
```{r}
## Take the annotation for the majority of the clusters as the final label
pbmc_rna$major_label <- NA

for (i in 1:17) {
  pbmc_rna$major_label[pbmc_rna$rna_clusters_r0.8==i] <- df_count[df_count$rna_clusters_r0.8==i,"predicted.id"][[1]]
}
```

```{r}
pbmc_rna$final_label <- pbmc_rna$major_label
pbmc_rna$final_label[Cells(pbmc_rna) %in% cell_inter] <- "B intermediate"
pbmc_rna$final_label[Cells(pbmc_rna) %in% cell_memory] <- "B memory"
pbmc_rna$final_label[Cells(pbmc_rna) %in% cell_cd8tem1] <- "CD8 TEM_1"
pbmc_rna$final_label[Cells(pbmc_rna) %in% cell_mait] <- "MAIT"
pbmc_rna$final_label[Cells(pbmc_rna) %in% cell_cd8tem2] <- "CD8 TEM_2"
pbmc_rna$final_label[Cells(pbmc_rna) %in% cell_gdt] <- "gdT"

pbmc_rna$final_label[pbmc_rna$rna_clusters_r0.8==2 & pbmc_rna$major_label=="CD4 Naive"] <- "CD4 Naive_1"
pbmc_rna$final_label[pbmc_rna$rna_clusters_r0.8==16 & pbmc_rna$major_label=="CD4 Naive"] <- "CD4 Naive_2"

pbmc_rna$final_label[pbmc_rna$rna_clusters_r0.8==3 & pbmc_rna$major_label=="CD8 Naive"] <- "CD8 Naive_1"
pbmc_rna$final_label[pbmc_rna$rna_clusters_r0.8==11 & pbmc_rna$major_label=="CD8 Naive"] <- "CD8 Naive_2"

pbmc_rna$final_label[pbmc_rna$rna_clusters_r0.8==12 & pbmc_rna$major_label=="CD8 TEM"] <- "CD8 TEM_1"
pbmc_rna$final_label[pbmc_rna$rna_clusters_r0.8==13 & pbmc_rna$major_label=="CD8 TEM"] <- "CD8 TEM_2"

df_count <- dplyr::rename(dplyr::count(pbmc_rna[[c("rna_clusters_r0.8","final_label")]], rna_clusters_r0.8, final_label), Freq = n) 
df_count <- dplyr::arrange(df_count, rna_clusters_r0.8, -Freq)
df_count

pbmc_rna$final_label <- as.factor(pbmc_rna$final_label)
Idents(pbmc_rna) <-"final_label"

table(pbmc_rna$final_label)
```
```{r, fig.keep='all'}
plot16 <- DimPlot(pbmc_rna, reduction = "umap",label = T, repel=T, group.by = "final_label",cols="polychrome")  + ggtitle("Curated labels") 
plot17 <- DimPlot(pbmc_rna, reduction = "umap",label = T, repel=T, group.by = "rna_clusters_r0.8",cols="polychrome")  + ggtitle("Clustering") 
plot18 <- DimPlot(pbmc_rna, reduction = "umap",label = T, repel=T, group.by = "predicted.id",cols="polychrome")  + ggtitle("transfered labels") 
plot16
plot17
plot18
```
```{r, eval=FALSE}
pdf(file = paste0(proj_dir,"PBMC_multi-omics_10X/rna/final_UMAP.pdf"), width = 10, height = 6)
plot16
plot17
plot18
dev.off()
```

```{r, fig.keep='all',message=FALSE,warning = FALSE}
Idents(pbmc_rna) <-"final_label"
pbmc_rna$seurat_clusters <- pbmc_rna$final_label
pbmc.markers <- FindAllMarkers(pbmc_rna, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
pbmc.markers %>%
    group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC)

pbmc.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) %>% arrange(cluster) -> top10

set1 <- c("B intermediate","B memory","CD14 Mono","B naive")
set2 <- c("CD16 Mono","CD4 Naive_1","CD4 Naive_2","CD4 TCM")
set3 <- c("NK","CD8 Naive_1","CD8 Naive_2","CD8 TEM_1","CD8 TEM_2")

pbmc1 <- subset(x=pbmc_rna,subset=final_label %in% set1)
pbmc2 <- subset(x=pbmc_rna,subset=final_label %in% set2)
pbmc3 <- subset(x=pbmc_rna,subset=final_label %in% set3)
DoHeatmap(pbmc1, features = top10$gene)
DoHeatmap(pbmc2, features = top10$gene) 
DoHeatmap(pbmc3, features = top10$gene) 
```

```{r}
# have a list of mutual exclusive marker genes used for manual checking?
genes <- c('TNFRSF13B', 'LINC01857', 'CD79B','SSPN', 'TEX9','IL4R','CXCR4', 'BTG1',"S100A9","CTSS", "S100A8","CDKN1C", "FCGR3A", "PTPRC","TCF7", "CD4","FHIT", "LEF1", "MAL","TMSB10","ITGB1","TRAC")
Idents(pbmc_rna) <- pbmc_rna$final_label
DotPlot(object = pbmc_rna, features = genes)
```

## Save the object
```{r}
saveRDS(pbmc_rna, file = "pbmc_rna3_with_curated_annotation_leiden.rds")
write.table(pbmc_rna[[c("final_label","atac_barcode")]], file='PBMC_multiomics_10X_predicted_cell_type_from_RNA_curated_leiden_PCA11.txt', sep='\t', quote = FALSE)
```

# check PCA N20
```{r}
DimPlot(pbmc_rna, reduction = "umap", group.by="SCT_snnPCAn20_res.0.8",cols="polychrome")
```
```{r}
## Check the consensus between clustering and label transfer
df_count <- dplyr::rename(dplyr::count(pbmc_rna[[c("SCT_snnPCAn20_res.0.8","predicted.id")]], SCT_snnPCAn20_res.0.8, predicted.id), Freq = n) 
df_count <- dplyr::arrange(df_count, SCT_snnPCAn20_res.0.8, -Freq)
df_count
```


